---
title: "HW 2: Exploration "
format: 
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
theme: Litera
editor: visual
---

Load Libraries

```{r}
#| message: false
#| warning: false

#libraries: 
library(tidyverse)   # dplyr, purrr, etc.
library(janitor)     # for clean_names()
library(lubridate)   # for mdy(), etc
library(here)        # for here()
library(sf)          # spatial data
library(doBy)        # for summaryBy()
```

## AQI

### Median AQI

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

# data ranges from 2000 - 2024
years <- 2000:2024

# -------- S E T  U P  L O O P  --------
median_aqi_list <- map(years, ~ {
  
      # build the file path
  aqi_file_path <- here("data/aqi_daily", paste0("aqidaily_", .x, ".csv"))
  
# -------- R E A D  I N  D A T A --------
  read_csv(aqi_file_path) %>%
    clean_names() %>%
    
# -------- T I D Y--------
      #transform date from a character to a date
    mutate(date = mdy(date)) %>%
    
      #extract year
    mutate(year = year(date)) %>% 

      #calculate median AQI
    group_by(year) %>%
    summarize(
      median_aqi = median(overall_aqi_value, na.rm = TRUE),
      .groups = "drop")
    
})

# combine into one data frame
median_aqi_df <- bind_rows(median_aqi_list)




  
```

### Visualize

```{r}
ggplot(median_aqi_df, aes(x = year, y = median_aqi)) +
  geom_line() + 
  geom_point() +
labs(x = "Year", y = "Mean AQI") +
  coord_fixed(ratio = 1/2) +
  theme_minimal()
```

Notes: fix scaling, aspect, overlay EPA's air quality standards as a reference line.

Events: California Diesel Risk Reduction Program (2000), 2005: The SCAQMD developed Clean Fleet rules to reduce diesel exhaust, decomissioning coal plants (?), 2003, SCAQMD implemented "Check Before You Burn, 2020 spike due to wildfire (?)

### Main Pollutant in AQI

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

# -------- S E T  U P  L O O P  --------
common_pollutant_list <- map(years, ~ {
  
  # file path
  aqi_file_path <- here("data/aqi_daily", paste0("aqidaily_", .x, ".csv"))
  
  # -------- R E A D  I N  D A T A --------
  read_csv(aqi_file_path) %>%
    clean_names() %>%
    
    # character to date
    mutate(date = mdy(date)) %>%
    
    # extract year
    mutate(year = year(date)) %>%
    
    select(year, main_pollutant)

})

# Combine all common pollutant calculations into one data frame
common_pollutant_all <- bind_rows(common_pollutant_list)


```


### Visualize 1.0

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: true

ggplot(common_pollutant_all, aes(x = main_pollutant)) +
  geom_bar() + 
  coord_flip()

```

Id love for this to be circles with size showing count...


```{r}
# ---------- SUMMARIZE BY YEAR ---------

# count occurrences of pollutants for the year
top_pollutants <- common_pollutant_all %>% 
  group_by(year, main_pollutant) %>%
  summarize(count = n(), .groups = "drop") %>% 

  group_by(main_pollutant) %>%
  summarize(avg_count = mean(count), .groups = "drop")



```

### Visualize 2.0
```{r}
library(dplyr)
library(packcircles)
library(ggplot2)

# -------- STEP 1:Compute circle layout----------

# circleProgressiveLayoutcalculates the positions and sizes of circles to create a non-overlapping
circle_layout <- circleProgressiveLayout(top_pollutants$avg_count, sizetype = "area")

# Add a small gap between circles
circle_layout$radius <- circle_layout$radius * 0.95

# Add coordinates back to the pollutant data
circle_data <- cbind(top_pollutants, circle_layout)

# create vertices for circle polygons
circle_vertices <- circleLayoutVertices(circle_layout, npoints = 50)

# -------- STEP 2: ADD NESTED CIRCLES ----------
# Generate radius, x, and y coordinates for nested circles
circle_layout_border <- circle_layout
circle_layout_border$radius <- circle_layout$radius * 0.95

# Create vertices for nested circles
circle_vertices_border <- circleLayoutVertices(circle_layout_border, npoints = 50)

# # -------- STEP 3: JOIN TABLES TO ADD LABELS ----------
# 
# Combine the pollutant data with circles (center coordinates and radius)
circle_labels <- cbind(top_pollutants, circle_layout)


```

```{r}
# -------- STEP 4: BASE PLOT(s) ----------

#plot the circles
gg_circles <- ggplot() +
  geom_polygon(data = circle_vertices, 
               aes(x = x, y = y, group = id, fill = as.factor(id)), 
               color = "black", alpha = 0.6) +
  theme_void() +
  coord_equal() +
  labs(fill = "Pollutant")


# add nested circles
  
gg_circles_nested <- gg_circles +
  geom_polygon(data = circle_vertices_border,
               aes(x = x, y = y, group = id, fill = as.factor(id)),
               colour = "black", alpha = 0.3) +
  theme_void()

gg_circles_nested 



```


```{r}


# custom red colors:
custom_red_palette <- c("#FFCCCC", "#FF9999", "#FF6666", "#FF3333", "#CC0000")


gg_circles_with_labels <- gg_circles_nested +
  geom_text(data = circle_labels,
            aes(x = x, y = y, size = avg_count, label = main_pollutant), # Add labels to the centers of the circles
            # fontface = "bold",
            color = "black") +
  scale_fill_manual(values = custom_red_palette) +  # Apply custom red palette
  scale_size_continuous(range = c(1.5, 6)) +

  theme_void() +  # Minimalist theme
  theme(legend.position = "none"
  ) +
  coord_equal()  # Ensure circles are proportional

gg_circles_with_labels



```

```{r}
# Calculate Percentages:
# Add a new column with the percentage of the average frequency of pollutants
circle_labels$percent <- round(
  circle_labels$avg_count / sum(circle_labels$avg_count) * 100
)

# custom red colors:
custom_red_palette <- c("#FFCCCC", "#FF9999", "#FF6666", "#FF3333", "#CC0000")

# Add percentage labels inside circles
gg_circles_nested +
  geom_text(
    data = circle_labels,
    aes(x = x, y = y, size = avg_count, label = paste0(percent, "%")),  # Show percentages only
    show.legend = FALSE  # Ensure text size doesn't appear in legend
  ) +
  scale_size_continuous(range = c(2, 10)) +  # Adjust text size scaling
  scale_fill_manual(
    values = custom_red_palette,  # Use custom color palette
    labels = circle_labels$main_pollutant  # Map pollutant names to legend
  ) +
  guides(fill = guide_legend(
    nrow = 3,
    direction = "horizontal",
    title = NULL,
    label.theme = element_text(size = 7, margin = margin(l = 0.5)),
    keywidth = unit(7, "pt"),
    keyheight = unit(7, "pt"))
  ) +
  theme_void() +  # Clean theme
  theme(
    legend.position = c(1.05, 0.15)  # Adjust legend position
  )

```



# Individual Pollutants

## PM 2.5 Single Pollutant

After exploring the data, I dont think I will use these datasets.

### Tidy

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

PM2_5_df <- read.csv(here("data/pm_annual.csv")) %>% 
  clean_names()

PM2_5 <- PM2_5_df %>%
  
  # transform date from a character to a date
  mutate(date = mdy(date)) %>%
  
  # create new columns for month, day, and year using lubridate
  mutate(
    month = month(date),  
    day   = day(date),    
    year  = year(date)    
  ) %>%
  
  # select only the columns we need
  select(daily_mean_pm2_5_concentration, units, year, month) 

# -------- F I N D   A N N U A L  M E A N --------  
   annual_PM2_5 <- PM2_5  %>%  
  group_by(year) %>%
    summarize(
      units = first(units),
      yearly_mean_pm2_5 = mean(daily_mean_pm2_5_concentration, na.rm = TRUE),
      .groups = "drop"
    )

```

### Visualize by Year

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

ggplot(annual_PM2_5, aes(x = year, y = yearly_mean_pm2_5)) +
  geom_line() + 
  geom_point() +
  labs(x = "Year", y = "Mean PM2.5 (µg/m³)") +
  theme_bw()



ggplot(annual_PM2_5, aes(x = year, y = yearly_mean_pm2_5)) +
  geom_line() + 
  geom_point() +
  labs(x = "Year", y = "Mean PM2.5 (µg/m³)") +
  theme_bw() +
  coord_fixed(ratio = 1/2) +
  scale_y_continuous(breaks = seq(0, 21, by = 5)) 


```

### Visualize by Month

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false


monthly_PM2_5 <- PM2_5 %>% 
   group_by(month) %>%
    summarize(
      units = first(units),
      monthly_mean_pm2_5 = mean(daily_mean_pm2_5_concentration, na.rm = TRUE),
      .groups = "drop"
    )
  
```

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

ggplot(monthly_PM2_5, aes(x = month, y = monthly_mean_pm2_5)) +
  geom_bar(stat = "identity") +
  labs(x = "Month (2000-2024)", y = "Mean PM2.5 (µg/m³)") +
  coord_fixed(ratio = 0.5) +
  theme_bw()
```

## Ozone

### Tidy

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

#read in data and clean names
annual_ozone <- read.csv(here("data/ozone_annual.csv")) %>% 
  clean_names()

annual_ozone <- annual_ozone %>% 
# transform date from a character to a date
  mutate(date = mdy(date)) %>%
  
# create new columns for month, day, and year using lubridate
  mutate(
    month = month(date),  
    day   = day(date),    
    year  = year(date)    
  ) %>% 
  
# group by year to show count of how many days per year exceed 0.070 ppm
  group_by(year) %>%
  summarize(days_over_070 = sum(daily_max_8_hour_ozone_concentration > 0.070, na.rm = TRUE))

```

### Visualize

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

# make a bar graph 
ggplot(annual_ozone, aes(x = year, y = days_over_070)) +
  geom_col()

#this is cool but id rather have it in buckets of unhealthy for sensi, unhealthy, and v unhealthy - stacked bar/area
```

# Census and Socioeconomic Data

## Enviroscreen (tabular) - Tidy

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

# read in enviroscreen data and clean names
ES_full <- read.csv(here("data/enviroscreen/enviroscreen_csv/enviroscreen_data.csv")) %>% 
  clean_names()

# ----- TIDY DATA ------
es_df <- ES_full %>% 
  # filter to only show LA County
  filter(california_county == "Los Angeles") %>% 
  # select for the relevant columns
  select(census_tract, 
         california_county, 
         total_population,
         zip,
         approximate_location,
         longitude,
         latitude,
         ozone,
         pm2_5,
         diesel_pm,
         tox_release,
         traffic,
         asthma,
         poverty,
)

# filter dataset for Los Angeles County
top_10_pm2_5  <- es_df %>% 
  filter(california_county == "Los Angeles")

top_10_pm2_5 <- top_10_pm2_5  %>%
  group_by(zip) %>% 
      # calculate the mean PM2.5 for each zip
  mutate(mean_pm2_5 = mean(pm2_5, na.rm = TRUE)) %>% 
  ungroup() %>%
      # arrange by the mean
  arrange(desc(mean_pm2_5)) %>% 
      # keep one row per zip
  distinct(zip, .keep_all = TRUE) %>% 
      # select the top 10 zip codes
  slice(1:10) 


```

Top 10 Zip Codes w highest mean PM 2.5:

-   91335 Reseda
-   91711 Claremont
-   91767 Pomona
-   90012 Downtown Los Angeles
-   91316 Encino
-   91766 Pomona
-   90031 Lincoln Heights --\> **MY HOUSE!**
-   90071 Bunker Hill/DTLA
-   90033 Boyle Heights
-   91406 Van Nuys

## Enviroscreen (spatial) - Tidy

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

# --------------------Spatial Data-----------------------
# bring in enviroscreen shapefile
enviroscreen_sf <- read_sf(here("data/enviroscreen/enviroscreen_shapefiles/CES4_final_shapefile.shp")) %>% 
  clean_names() 

# ----- TIDY DATA ------
enviroscreen_sf <- enviroscreen_sf %>% 
  # filter to only show LA County
  filter(county == "Los Angeles") %>% 
  # select for the relevant columns
  select(tract, 
         zip,
         approx_loc,
         ozone,
         pm2_5,
         pm2_5_p,
         diesel_pm,
         traffic,
         asthma,
         poverty,
         geometry,  
         county
)



```

## Tidy Census

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

library(tidycensus)

### Step 1: Census data filtered to Los Angeles
# all of LA county
la_census <- tidycensus::get_acs(
  state = "CA",
  county = "Los Angeles",
  geography = "tract",
  variables = "B25004_001", ## Vacancy Status
  geometry = TRUE,
  year = 2022
)

# filtered to exclude unincorporated areas and islands
la_census_filter <- la_census %>%
  mutate(NAME = gsub(", Los Angeles County, California", # elements that you want to remove
                     "", # replace with blank
                     NAME)) %>%
  mutate(NAME = gsub("Census Tract ", # elements that you want to remove
                     "", # replace with blank
                     NAME)) %>%
  filter(GEOID != "06037599100") %>% # islands
  filter(GEOID != "06037599000") %>% # islands
  filter(GEOID != "06037980003") %>%
  filter(GEOID != "06037980004") %>%
  filter(!(NAME >= 9000 & NAME <= 9800))

### Step 2: Enviroscreen data filtered to Los Angeles
# read in and filter out areas we know we don't want
enviroscreen_sf <- enviroscreen_sf %>%
  janitor::clean_names() %>%
  filter(county == "Los Angeles") %>%
  filter(tract != 6037599100) %>% # islands
  filter(tract != 6037599000) %>% # islands
  filter(tract != 6037980003) %>% # unincorporated
  filter(tract != 6037930101) %>% # forests, unincorporated areas, etc
  filter(tract != 6037930301) %>%
  filter(tract != 6037930200) %>%
  filter(tract != 6037920303)

```

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

### Step 3: join tidy census and enviroscreen
# set crs to be the same as la census tracts
enviroscreen_sf <- st_transform(enviroscreen_sf, st_crs(la_census_filter))

# filter out northern areas so we just have LA itself left
final_enviroscreen <- st_filter(enviroscreen_sf, la_census_filter)
```

### Visualize

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: true

# plot PM 2.5 per census tract
ggplot() +
  geom_sf(data = final_enviroscreen, aes(fill = pm2_5_p), color = 'black', size = 1) +
  scale_fill_distiller(palette = "OrRd", direction = 1) +
  theme_minimal() +
  labs(fill = "PM2.5 Percentile") +
  theme(
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.text = element_blank(),
    axis.title = element_blank())+
  theme_void()
```

Using percentile PM2.5 as opposed to raw PM 2.5 shows more of an EJ POV, as the percentile is a relative ranking of PM2.5 concentrations compared to other census tracts statewide. (ex: A percentile of 90 means the area has higher PM2.5 pollution than 90% of other census tracts in California.)

# Pollution Sources

## Point Sources

### Tidy

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

point_sources <- read.csv(here("data/sources/facility_point_sources_2020.csv")) %>% 
  clean_names()

point_sources <- point_sources %>% 
  filter(pollutant %in% c("PM2.5 Primary (Filt + Cond)", "PM2.5 Filterable"))

# split the lat_lon into two separate columns to create spatial object

point_sources <- point_sources %>%
    # Remove brackets from the lat_long column
    mutate(lat_long = gsub("\\[|\\]", "", lat_lon)) %>%
    # Separate the lat_long column into two columns: longitude and latitude
    separate(lat_long, into = c("longitude", "latitude"), sep = ",") %>%
    # Convert to numeric
    mutate(
        longitude = as.numeric(longitude),
        latitude = as.numeric(latitude)
    )

# Convert to spatial object 
point_sources_sf <- st_as_sf(
  point_sources,
  coords = c("longitude", "latitude"),  # REMINDER: Longitude comes first!
  crs = 4326  # WGS84 CRS 
)

# Transform to match CRS of `final_enviroscreen`
point_sources_sf <- st_transform(point_sources_sf, st_crs(final_enviroscreen))

# filter out northern areas so we just have LA itself left
point_sources_sf <- st_filter(point_sources_sf, final_enviroscreen)

```

### Visualize Spatially

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: false

ggplot() +
    geom_sf(data = point_sources_sf, aes(color = pollutant)) +
    theme_minimal() +
    labs(title = "Facility Point Sources", color = "Pollutant")

```

There are just so many point sources, so I'm going to try mapping just the top what it I mapped just the top 50 emitters.

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

# only show emissions over 6.1 tons

#convert emissions_tons to numeric: 
point_sources_sf <- point_sources_sf %>%
  mutate(emissions_tons = as.numeric(emissions_tons))

# filter for top 50 emitters
top_point_sources <- point_sources_sf %>% 
  filter(emissions_tons > 6.1)

# filter for major sources only
major_point_sources <- point_sources_sf %>% 
  filter(emissions_tons >= 100)
```

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: true

ggplot() +
  # LA map with PM2.5 fill
  geom_sf(data = final_enviroscreen, aes(fill = pm2_5_p), color = 'black', size = 0.2, alpha = 1) +
  
  # Add facility points with customized colors
  geom_sf(data = top_point_sources, aes(color = pollutant), size = 2, alpha = 1) +
  
  # Fill scale for PM2.5
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "gray90") +
  
  # Color scale for pollutant types
  scale_color_manual(
    values = c(
      "PM2.5 Primary (Filt + Cond)" = "cyan",
      "PM2.5 Filterable" = "cyan4"
    )
  ) +
  
  # Minimalist theme
  theme_void()


```

### Visualize by Source Type

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: true

point_sources_plot <- ggplot(top_point_sources, aes(x = reorder(facility_type, emissions_tons), y = emissions_tons)) +
  geom_col() +  
  coord_flip() +  
  labs(
    title = "PM2.5 Emissions - Point Sources (Facilities)",
    x = "Facility Type",
    y = "Emissions (Tons)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 6)
  )

point_sources_plot


```

## Non-point Sources

# Tidy

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

nonpoint_sources <- read.csv(here("data/sources/nonpoint_sources_2020.csv")) %>% 
  clean_names()

nonpoint_sources <- nonpoint_sources %>% 
  filter(pollutant %in% c("PM2.5 Primary (Filt + Cond)", "PM2.5 Filterable")) %>% 
  
#convert emissions_tons to numeric
  mutate(emissions_tons = as.numeric(emissions_tons)) %>% 
  drop_na()

# filter for top 50 emitters
top_nonpoint_sources <- nonpoint_sources %>% 
  filter(emissions_tons > 6.1)

# filter for major sources only
major_nonpoint_sources <- nonpoint_sources %>% 
  filter(emissions_tons >= 100)
  
```

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| output: true

nonpoint_sources_plot <- ggplot(major_nonpoint_sources, aes(x = reorder(scc_level_2, emissions_tons), y = emissions_tons)) +
  geom_col() +  
  coord_flip() +  
  labs(
    title = "PM2.5 Emissions - Nonpoint",
    x = "scc_level_2",
    y = "Emissions (Tons)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 6)
  )

nonpoint_sources_plot

```
